---
title: "Multivariate Data Visualization in R"
author: "Shrihari Kakakoti"
date: "2025-02-25"
output: html_document
---

# Introduction

This project explores multivariate data visualization techniques in R using a dataset of personal contacts. The aim is to highlight how data with multiple variables can be effectively represented and analyzed visually.

# Background

Multivariate data visualization helps in uncovering patterns, correlations, and insights in high-dimensional datasets. In this case study, we visualize attributes such as maximum temperature, weight, height, gender, year of acquaintance, and perceived company quality.

## Ask

How can we visually represent multivariate data to better understand relationships between variables and identify patterns?

## Prepare

We use a sample dataset containing the following variables:
- Contact
- Maxtemp
- Weight
- Height
- Year
- Gender
- Company

### Data Preparation

```{r load-libraries, message=FALSE}
library(ggplot2)
library(readxl)
library(plot3D)
library(GGally)
library(aplpack)
```

```{r load-data}
mydata <- read_excel("multivariate.xlsx")
head(mydata)
```

## Process

### Analyse

```{r vis1, echo=TRUE, fig.cap="2D Plot: Weight vs Height sized by Max Temperature"}
ggplot(mydata, aes(Weight, Height, size = Maxtemp)) + 
  geom_point(alpha = 0.6) + 
  labs(title = "Friends Contact Data: Weight, Height, Maxtemp")
```

```{r vis2, echo=TRUE, fig.cap="2D Plot: Weight vs Height colored by Company"}
ggplot(mydata, aes(Weight, Height, colour = Company)) + 
  geom_point(alpha = 0.6) + 
  labs(title = "Friends Contact Data: Weight, Height, Company")
```

```{r vis3, echo=TRUE, fig.cap="2D Plot: Weight vs Height colored by Company and sized by Max Temperature"}
ggplot(mydata, aes(Weight, Height, colour = Company, size = Maxtemp)) + 
  geom_point(alpha = 0.6) + 
  labs(title = "Friends Contact Data: Weight, Height, Maxtemp, Company")
```

```{r vis4, echo=TRUE, fig.cap="3D Plot: Weight vs Height vs Max Temperature"}
scatter3D(x = mydata$Weight, y = mydata$Height, z = mydata$Maxtemp, 
          bty = "f", colkey = FALSE, main = "Weight, Height, Maxtemp")
```

```{r vis5, echo=TRUE, fig.cap="Parallel Coordinates Plot by Company"}
ggparcoord(data = mydata, columns = 2:6, groupColumn = "Company")
```

```{r vis6, echo=TRUE, fig.cap="Chernoff Faces"}
faces(mydata[,2:5], face.type=1, scale=TRUE, main="Friends Contact Data Set", 
      labels=mydata$Contact, print.info=TRUE)
```

## Share

All visualizations above help communicate patterns and relationships within the data. Each image is generated as a `.png` when knitting this document.

## Act

### Conclusion

Multivariate visualization provides a powerful way to explore and understand complex datasets. Using R libraries like `ggplot2`, `plot3D`, `GGally`, and `aplpack`, we can create engaging and insightful representations.

### Recommendations

- Use parallel coordinates for comparing multiple variables across different categories.
- Use Chernoff faces sparingly and only when interpretability is understood by the audience.
- Complement these visualizations with statistical summaries for deeper analysis.
